Generalized Plackett-Luce Likelihoods

### 2023-05-08 inst/jss_4382_rebuttal.tex

Short story: I have accommodated all the comments with rewording and
clarification.

\subsection*{Reviewer's comments, 15 April 2023}

Dear Robin K.S. Hankin,

We have reached a decision regarding your submission to Journal of
Statistical Software, \verb+#4832+, ``Generalized Plackett-Luce
Likelihoods".

@@

1.1 Recommendation

The manuscript may be publishable in the journal after revisions have been
made.

1.2 General comments

G1 The introduction should provide an overview of the contents of the
manuscript. \\  \textcolor{blue}{New sentence added to manuscript:\\ \\
Here I consider such generalized Plackett-Luce likelihood functions,
and give an exact analysis of several simple cases.  I then show how
this class of likelihood functions may be applied to a range of
inference problems involving order statistics.  Illustrative examples,
drawn Formula 1 motor racing, and track-and-field athletics, are
given.}\\ \\

In addition, the section headlines could be made more
informative. Currently, they are:

2. ``Some simple cases"
3. ``Mann-Whitney test"
4. ``Multiple competitors: javelin"
5. ``Formula 1 motor racing: the constructors' championship"

Maybe the headlines could provide more of a summary of what the
section contains.  \textcolor{blue}{The headings are indeed somewhat
  uninformative and are now more in line with the remainder of the
  manuscript.  I have changed what was section 4 to a {\em sub}section
  with heading ``A generalization of the Mann-Whitney test using
  generalized Plackett-Luce likelihood'' which I think is in keeping
  with this comment.  The headings are now
  \begin{itemize}
  \item Exact analytical solution for some simple generalized
    Plackett-Luce likelihood functions
  \item An alternative to the Mann-Whitney test using generalized
    Plackett-Luce likelihood
  \item A generalization of the Mann-Whitney test using
    generalized Plackett-Luce likelihood (now a subsection)
  \item Formula 1 motor racing: the constructors' championship
\end{itemize}}
\rule{0mm}{10mm}

G2 There should be a comparison with other software.  Although the
author states in the introduction that with hyper2 ``It becomes
possible to analyse a wider range of likelihood functions than
standard Plackett-Luce", it is not clear to me which of the features
described later on are unique to the \verb+hyper2+ package and which
can also be obtained from other packages (like, e.g., PlackettLuce).

\textcolor{blue}{Although the hyper2 package has functionality not
  present in other implementations, the current manuscript focuses on
  the new generalization represented in the package by the hyper3
  class (that is, generalized Plackett-Luce likelihoods for order
  statistics with clones of identical Bradle-Terry strength).  Turner
  et. al. (2020), which I have cited, does a good job of setting out
  the functionality of hyper2 not present in other packages.  I think
  that an extended discussion of hyper2 unique functionality would be
  a distraction from the focus of the manuscript.  But I agree that a
  short statement documenting the unique aspects of the hyper2 package
  would be useful to JSS readers, and I have added a short paragraph
  to this effect.  Actually (in the spirit of full disclosure), I have
  added quite a lot of machinery to the package since Turner's 2020
  review and this gives me a welcome opportunity to mention some of
  it.}

G3 The manuscript should comment on the timing of the functions. For the
simple example on P. 4, the maxp(H) call takes more than 8 seconds on
my computer. This makes me wonder how timings would be for realistic
examples.

\textcolor{blue}{Any optimization technique becomes slow when used
  over a high-dimensional space and {\tt maxp()} is no exception.  The
  examples chosen were well within its capabilities.  In addition, the
  default implementation of {\tt maxp()} in the package uses a *very*
  conservative set of assumptions, for example the default is to use
  $n=10$ slightly differing start points and perform (by default)
  $n-10$ independent optimizations, to check for
  convergence.  I have come to realise that this is absurdly
  over-engineered for the examples considered in the manuscript.  If
  one is sure that the convergence is guaranteed (as it is in the
  examples considered here), $n=1$ is perfectly adequate.  For
  example, using the case given on P4:
  {\tt\\ \\
> (system.time(mH <- maxp(H,n=1)))\\
   user  system elapsed \\
  0.355   0.003   0.357 \\
> maxp(H,n=1)\\
         a          b          c \\
0.21324090 0.08724824 0.69951086 \\
  }\\ \\
  The latest package version on github includes a discussion of this
  issue and I have added a brief discussion (``function {\tt maxp()}
  uses standard optimization techniques to locate the evaluate; it has
  access to first derivatives of the log-likelihood and as such has
  rapid convergence, if its objective function is reasonably
  smooth").}\\ \\

1.3 Specific comments\\ S1 I'm not claiming any expertise in English
language, but occasionally, I found the manuscript difficult to
understand. For example, in the abstract it says: "The package
. . . furnishes idiom" Couldn't that simply be: "The package
. . . provides functionality"?

\textcolor{blue}{The original motivation for the package occurred when
  I had a large complicated set of Bradley-Terry observations which I
  wanted to translate into a likelihood function.  The R code became
  extremely long and unweildy, and was almost impossible to check.  It
  also became very slow to run.  What I wanted was a nice way to enter
  this kind of observations, in such a way as to maintain both
  computational efficiency [hence the use of the {\tt STL} map class]
  {\bf and} human readability.  I think that the phrase ``furnishes
  idiom" expresses this reasonably clearly.  But on reflection it does
  sound a bit odd and I have rephrased the abstract accordingly.}\\ \\

On P. 13, it says "Here we use generalized Plackett-Luce to assess the
constructors' performance". Isn't there a "model" missing?
(Plackett-Luce is just a name and cannot be used).
\textcolor{blue}{The original text was too terse and I have
  reworded.}\\ \\

### 2023-05-08 inst/jss_hyper3.Rnw
typo
Acked-by: Robin Hankin <hankin.robin@gmail.com>
more work on issue #153

  The \code{hyper2} package furnishes extensive idiom to work with
  extensions of the Bradley-Terry probability model such as
  Plackett-Luce likelihood functions, team strengths, and reified
  entities (monsters).  Here, I present a generalization of
  \code{hyper2} in which multiple entities are constrained to have
  identical Bradley-Terry strengths.  A new \proglang{S3} class
  \code{hyper3}, along with associated methods, is motivated and
  introduced.  Three datasets are analysed, each analysis furnishing
  new insight, and each highlighting different capabilities of the
  package.
}
\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable. Further, the
introduction of reified entities allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.

@@

@@

@@

\code{Package helper functions}

@@

\section{Some simple cases}

@@

\begin{figure}[htbp]
  \begin{center}
<<figabbabb,echo=FALSE,fig=TRUE>>=
f_aab <- function(a){a^2 / (1 + a)}
f_aba <- function(a){a * (1 - a) / (1 + a)^2}
f_baa <- function(a){(1 - a) / (1 + a)}

p <- function(f, ...){points(a, f(a) / max(f(a)), ...)}
plot(0:1, 0:1, xlab = "p(a)", ylab = "Likelihood", type = "n")
p(f_aab, type = "l", col = "black")
p(f_aba, type = "l", col = "red")
p(f_baa, type = "l", col = "blue")
legend(x=0.85, y=0.7, col = c("black", "red", "blue"), lty=1, pch=NA,
       legend=c("AAB", "ABA", "BAA"))
abline(h = exp(-2), lty = 2)
@
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.}
\end{center}
\end{figure}

@@

\section{Mann-Whitney test}

@@

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,echo=FALSE,fig=TRUE>>=
a <- seq(from = 0.02, to = 0.8, len = 40)
L <- sapply(a, function(p){loglik(p, Hxy)})
plot(a, L - max(L), type = 'b')
abline(h = c(0, -2))
abline(v = c(0.24))
abline(v=c(0.5), lty=2)
@
\caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
strength $p_a$ of permeability at term.  The evaluate of 0.24 is
shown; and the two-units-of support credible interval, which does not
exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
\end{center}
\end{figure}

\section{Multiple competitors: Javelin}

@@

\begin{figure}[htbp]
\begin{center}
<<plottheloglikcont,echo=FALSE,fig=TRUE>>=
colnames(M) <-  as.character(Ves)
plot(t(M), type = "b")
abline(h = c(0, -2))
abline(v = 0, lty = 2)
abline(v = log(0.32062833 / 0.11402735)) # these from mp
@
\caption{Profile \label{profliklogcont} likelihood for log-contrast
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
line, and two-units-of-support limit indicated with horizontal lines;
thus the null is not rejected.}
\end{center}
\end{figure}

@@

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In Formula 1 Grand Prix, the
constructors' championship takes place between {\em manufacturers} of
racing cars (compare the drivers' championship, which is between
drivers).  In this analysis, the constructor is the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the constructors' performance.
The following table, included in the \code{hyper2} package as a
dataset, shows rankings for the first 9 venues of the 2021 season:

>>>>

  The \code{hyper2} package provides functionality to work with
  extensions of the Bradley-Terry probability model such as
  Plackett-Luce likelihood including team strengths and reified
  entities (monsters).  The package allows one to use relatively
  natural R idiom to manipulate such likelihood functions.  Here, I
  present a generalization of \code{hyper2} in which multiple entities
  are constrained to have identical Bradley-Terry strengths.  A new
  \proglang{S3} class \code{hyper3}, along with associated methods, is
  motivated and introduced.  Three datasets are analysed, each
  analysis furnishing new insight, and each highlighting different
  capabilities of the package.  } \Keywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney}

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  The \pkg{hyper2}
package has the ability to evaluate such likelihood functions at any
point in $S_n$, thereby admitting a wide range of non-standard nulls
such as order statistics on the $p_i$~\cite{hankin2020}.  It becomes
possible to analyse a wider range of likelihood functions than
standard Plackett-Luce~\citep{turner2020}.  For example, results
involving incomplete order statistics or teams are tractable.
Further, the introduction of reified entities (monsters) allows one to
consider draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.  Recent versions of the package include
experimental functionality (\code{cheering()}) to investigate the
relaxing of the assumption of conditional independence of the
forward-ranking process.

@@

Here I consider such generalized Plackett-Luce likelihood functions,
and give an exact analysis of several simple cases.  I then show how
this class of likelihood functions may be applied to a range of
inference problems involving order statistics.  Illustrative examples,
drawn Formula 1 motor racing, and track-and-field athletics, are
given.

@@

(function \code{maxp()} uses standard optimization techniques to
locate the evaluate; it has access to first derivatives of the
log-likelihood and as such has rapid convergence, if its objective
function is reasonably smooth").

@@

\subsection{Package helper functions}

@@

\section{Exact analytical solution for some simple generalized
Plackett-Luce likelihood functions}

@@

\begin{figure}[htbp]
  \begin{center}
<<figabbabb,echo=FALSE,fig=TRUE>>=
f_aab <- function(a){a^2 / (1 + a)}
f_aba <- function(a){a * (1 - a) / (1 + a)^2}
f_baa <- function(a){(1 - a) / (1 + a)}
p <- function(f, ...){
  a <- seq(from = 0, by = 0.005, to = 1)
  points(a, f(a) / max(f(a)), ...)
  }
plot(0:1, 0:1, xlab = expression(p[a]), ylab = "Likelihood", type = "n")
p(f_aab, type = "l", col = "black")
p(f_aba, type = "l", col = "red")
p(f_baa, type = "l", col = "blue")
text(0.8,0.8,"AAB")
text(0.8,0.5,"ABA",col="red")
text(0.8,0.15,"BAA",col="blue")
abline(h = exp(-2), lty = 2)
@
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.}
\end{center}
\end{figure}

@@

\section{An alternative to the Mann-Whitney test using generalized
Plackett-Luce likelihood}

@@

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,echo=FALSE,fig=TRUE,cache=TRUE>>=
a <- seq(from = 0.02, to = 0.8, len = 40)
L <- sapply(a, function(p){loglik(p, Hxy)})
plot(a, L - max(L), type = 'b')
abline(h = c(0, -2))
abline(v = c(0.24))
abline(v=c(0.5), lty=2)
@
\caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
strength $p_a$ of permeability at term.  The evaluate of 0.24 is
shown; and the two-units-of support credible interval, which does not
exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
\end{center}
\end{figure}

\subsection{A generalization of the Mann-Whitney test using
generalized Plackett-Luce likelihood}

@@

\begin{figure}[htbp]
\begin{center}
<<plottheloglikcont,echo=FALSE,fig=TRUE,cache=TRUE>>=
colnames(M) <-  as.character(Ves)
plot(t(M), type = "b")
abline(h = c(0, -2))
abline(v = 0, lty = 2)
abline(v = log(0.32062833 / 0.11402735)) # these from mp
@
\caption{Profile \label{profliklogcont} likelihood for log-contrast
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
line, and two-units-of-support limit indicated with horizontal lines;
thus the null is not rejected.}
\end{center}
\end{figure}

@@

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In Formula 1 Grand Prix, the
constructors' championship takes place between {\em manufacturers} of
racing cars (compare the drivers' championship, which is between
drivers).  In this analysis, the constructor is the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized a Plackett-Luce model to assess the constructors'
performance.  The following table, included in the \code{hyper2}
package as a dataset, shows rankings for the first 9 venues of the
2021 season:
